Map{R}educe in {MPI} for Large-scale Graph Algorithms

### 2010-04-28 paper/graph.tex
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@587 3fc257bb-8550-0410-81b3-a5c103da83f7

We begin with a MapReduce procedure for creating large, sparse,
randomized graphs, since they are the input for the algorithms
discussed below.  R-MAT graphs ~\cite{RMAT} are recursively generated
graphs with power-law degree distributions.  They are commonly used to
represent web and social networks.  The user specifices 6 parameters
which define the graph: the number of vertices $N$ and edges $M$, and
4 parameters $a$, $b$, $c$, $d$ which sum to 1.0 and are discussed
below.  The algorithm in Figure \ref{fig:rmat} generates $N_z = MN$
unique non-zero entries in a sparse $NxN$ matrix $A$, where each entry
$A_{ij}$ represents an edge between graph vertices $(V_i,V_j)$.

@@

$N_{\rm remain} = N_z$ \\
while $N_{\rm remain} > 0$: \\
\> 1 Map: \> Generate $N_{\rm remain}/P$ random edges on each processor \\
           \> \> output Key = $(V_i,V_j)$, Value = NULL \\
\> 1 Collate \\
\> 1 Reduce: \> Remove duplicate edges \\
              \> \> input Key = $(V_i,V_j)$, MultiValue = one or more NULLs \\
              \> \> output Key = $(V_i,V_j)$, Value = NULL \\
\> $N_{\rm remain} = N_z - N_{kv}$

@@

The map() will often generate some small number of duplicate edges.
The collate() and reduce() operations remove the duplicates.  The
entire map-collate-reduce sequence is repeated until the number of
resulting key/value pairs $N_{kv}$ equals $N_z$.  For reasonably
sparse graphs this typically takes only a few iterations.

@@

Assign random values to each vertex \\
Clone: convert edge key/values directly to key/multivalues \\
Create empty MapReduce MRv for maximal independent set vertices \\
\\
while $N_{edges} > 0$: \\
\> 1 Reduce: \> Determine winner/loser vertex of each edge \\
       \> \> \> input Key = $E_{ij}$, Multivalue = NULLS or LOSER \\
       \> \> \> if any value is LOSER, emit nothing \\
       \> \> \> $V_w$ = WINNER vertex, $V_l$ = LOSER vertex \\
       \> \> \> output Key = $V_w$, Value = $V_l$ WINNER \\
       \> \> \> output Key = $V_l$, Value = $V_w$ LOSER \\
\> 1 Collate: \> Vertex as key \\
\> 2 Reduce: \> Find WINNER vertices \\
       \> \> \> input Key = $V_i$, MultiValue = $V V V V ...$ \\
       \> \> \> if all $V$ are WINNERs: \\
      \> \> \> \> output to MRv, Key = $V_i$, Value = NULL \\
      \> \> \> \> for each $V$ in Multivalue: \\
     \> \> \> \> \> output Key = $V_j$, Value = $Vi$ LOSER \\
       \> \> \> else: \\
      \> \> \> \> for each $V$ in Multivalue: \\
     \> \> \> \> \> output Key = $V_j$, Value = $V_i$ \\
\> 2 Collate: \> Vertex as key \\
\> 3 Reduce: \> Find LOSER vertices, delete their edges \\
       \> \> \> input Key = $V_i$, MultiValue = $V V V V ...$ \\
       \> \> \> if any $V$ is LOSER: \\
      \> \> \> \> for each $V$ in Multivalue: \\
     \> \> \> \> \> output Key = $E_{ij}$, Value = LOSER \\
	\> \> \> else: \\
      \> \> \> \> for each $V$ in Multivalue: \\
     \> \> \> \> \> output Key = $E_{ij}$, Value = NULL \\
\> 3 Collate: \> Edge as key \\

@@

 \caption{MapReduce algorithm for finding a maximal independent set of
 graph vertices.}

@@

Before the iterations begin, a random value is assigned to each
vertex, which will be used to compare pairs of vertices.  The vertex
ID itself could be used as the random value, but since the vertices
eventually selected for the MIS depend on the randomization, this may
introduce an unwanted bias.  Instead a random number generator can be
used to assign a random value $R_i$ to each vertex (during a map()
operation), which is then carried along with the vertex ID through
each stage of the algorithm.  In the notation of Figure
\ref{fig:luby}, each $V_i$ is then really two quantities, a vertex ID
(1 to $N$) and $R_i$.  Alternatively, the $R_i$ can be used to relabel
the vertices in a random manner, assigning each a new, unique vertex
ID from 1 to $N$.  In this case, $V_i$ is simply the new vertex ID,
and the vertices in the final MIS could be remapped to recover their
original IDs.  These operations can be performed with a few extra
MapReduce stages, outside the iteration loop.

The pre-iteration clone() operation converts the initial list of edge
key/values one by one to key/multivalue pairs, with the edge as the
key, and NULL as the value.  This is so that the iteration can begin
with a reduce() operation, without need for a collate() operation.  A
second empty MapReduce object is also initialized, which will
accumulate MIS vertices as the iterations proceed.

At each iteration, the current set of edges are examined to identify
winning vertices.  There are removed and added to the MIS.  Vertices
adjacent to winners are flagged as losers, and all edges adjacent to
losers are flagged for removal at the start of the next iteration.
Thus each iteration removes edges from the initially full graph, and
adds vertices to the MIS.  When the graph is empty, the MIS is
complete.

The first reduce() operation flags the two vertices in each edge as a
potential winner or loser.  This is done by comparing the two random
values for the vertices (or the randomized vertex IDs as explained
above).  The first collate() operation thus produces a multivalue for
each vertex which contains all the vertices it shares an edge with and
associated winner/loser flags for $V_i$.  In the second reduce()
operation, if $V_i$ won the comparison with all its edge neighbors,
then it is a ``winner'' vertex and is emitted to the second MapReduce
object.  All vertices connected to the winner are emitted with a loser
flag.  If $V_i$ is not an overall winner, its edge vertices are
emitted without the loser flag.  The second collate() operation
collects this new information for each vertex.

In the third reduce() operation, if $V_i$ was flagged as a loser by
any edge neighbor who was a winner, then it becomes a loser.  All of
its edges are emitted with a loser flag.  If $V_i$ was not flagged as
a loser, its edges are emitted with no flag.  The final collate()
operation gathers this information for each edge; each edge will have
two values (from each vertex), which are either a loser flag or NULL.
At the first stage of the next iteration, any edge flagged as a loser
does not compare its two vertices; it is effectively deleted from the
graph.

The Luby iterations end when all edges have been removed from the
graph.  At this point, the second MapReduce object contains a complete
MIS of vertices.

>>>>

We begin with a MapReduce procedure for creating large, sparse,
randomized graphs, since they are the input for the algorithms
discussed below.  R-MAT graphs ~\cite{RMAT} are recursively generated
graphs with power-law degree distributions.  They are commonly used to
represent web and social networks.  The user specifices 6 parameters
which define the graph: the number of vertices $N$ and edges $M$, and
4 parameters $a$, $b$, $c$, $d$ which sum to 1.0 and are discussed
below.  The algorithm in Figure \ref{fig:rmat} generates $M$ unique
non-zero entries in a sparse $NxN$ matrix $A$, where each entry
$A_{ij}$ represents an edge between graph vertices $(V_i,V_j)$.

@@

$M_{\rm remain} = M$ \\
while $M_{\rm remain} > 0$: \\
\> 1 Map: \> Generate $M_{\rm remain}/P$ random edges on each processor \\
           \> \> output Key = $(V_i,V_j)$, Value = NULL \\
\> 1 Collate \\
\> 1 Reduce: \> Remove duplicate edges \\
              \> \> input Key = $(V_i,V_j)$, MultiValue = one or more NULLs \\
              \> \> output Key = $(V_i,V_j)$, Value = NULL \\
\> $M_{\rm remain} = M - N_{kv}$

@@

The map() will often generate some small number of duplicate edges.
The collate() and reduce() operations remove the duplicates.  The
entire map-collate-reduce sequence is repeated until the number of
resulting key/value pairs $N_{kv} = M$.  For reasonably sparse graphs
this typically takes only a few iterations.

@@

Map: assign random values to each vertex \\
Clone: convert edge key/values directly to key/multivalues \\
Create: empty MapReduce MRv for maximal independent set vertices \\
\\
while $N_{edges} > 0$: \\
\> 1 Reduce: \> Determine WINNER/LOSER vertex of each edge \\
       \> \> \> input Key = $E_{ij}$ with $V_i < V_j$, Multivalue = NULL or LOSER \\
       \> \> \> if either value is LOSER, emit nothing \\
       \> \> \> $V_w$ = WINNER vertex, $V_l$ = LOSER vertex \\
       \> \> \> output Key = $V_w$, Value = $V_l$ WINNER \\
       \> \> \> output Key = $V_l$, Value = $V_w$ LOSER \\
\> 1 Collate: \> Vertex as key \\
\> 2 Reduce: \> Find WINNER vertices \\
       \> \> \> input Key = $V_i$, MultiValue = $V V V V ...$ \\
       \> \> \> if all $V$ are WINNERs: \\
      \> \> \> \> for each $V$ in Multivalue: \\
     \> \> \> \> \> output Key = $V_j$, Value = $Vi$ LOSER \\
       \> \> \> else: \\
      \> \> \> \> for each $V$ in Multivalue: \\
     \> \> \> \> \> output Key = $V_j$, Value = $V_i$ \\
\> 2 Collate: \> Vertex as key \\
\> 3 Reduce: \> Find LOSER vertices \\
       \> \> \> input Key = $V_i$, MultiValue = $V V V V ...$ \\
       \> \> \> if any $V$ is LOSER: \\
      \> \> \> \> for each $V$ in Multivalue: \\
     \> \> \> \> \> output Key = $V_j$, Value = $Vi$ LOSER \\
       \> \> \> else: \\
      \> \> \> \> for each $V$ in Multivalue: \\
     \> \> \> \> \> output Key = $V_j$, Value = $V_i$ \\
\> 3 Collate: \> Vertex as key \\
\> 4 Reduce: \> Emit WINNER vertices and LOSER edges \\
       \> \> \> input Key = $V_i$, MultiValue = $V V V V ...$ \\
       \> \> \> if all $V$ are LOSERs: \\
      \> \> \> \> output to MRv, Key = $V_i$, Value = NULL \\
       \> \> \> for each $V$ in Multivalue: \\
      \> \> \> \> if $V$ is LOSER: \\
     \> \> \> \> \> output Key = $E_{ij}$ with $V_i < V_j$, Value = LOSER \\
      \> \> \> \> else: \\
     \> \> \> \> \> output Key = $E_{ij}$ with $V_i < V_j$, Value = NULL \\
\> 4 Collate: \> Edge as key \\

@@

 \caption{MapReduce algorithm for finding a maximal independent set of
 graph vertices via Luby's algorithm.}

@@

Before the iterations begin, a random value is assigned to each
vertex, which will be used to compare pairs of vertices.  The vertex
ID itself could be used as the random value, but since the vertices
eventually selected for the MIS depend on the randomization, this may
introduce an unwanted bias.  Instead a random number generator can be
used to assign a consistent random value $R_i$ to each vertex in each
edge (via a map() operation), which is then carried along with the
vertex ID through each stage of the algorithm.  In the notation of
Figure \ref{fig:luby}, each $V_i$ is then really two quantities, a
vertex ID (1 to $N$) and $R_i$.  Alternatively, the $R_i$ can be used
to relabel the vertices in a random manner, assigning each a new,
unique vertex ID from 1 to $N$.  In this case, $V_i$ is simply the new
vertex ID, and the vertices in the final MIS could be remapped to
recover their original IDs.  These operations could be performed with
a few extra MapReduce steps, outside the iteration loop.

The pre-iteration clone() operation converts the initial list of edge
key/values one by one to key/multivalue pairs, with the edge as the
key, and NULL as the value.  This is so the iterations can begin with
a reduce() operation, without need for a collate() operation.  A
second empty MapReduce object is also initialized, which will
accumulate MIS vertices as the iterations proceed.

At each iteration, the current set of edges are examined to identify
winning vertices.  Vertices adjacent to winners are flagged as losers,
and all edges adjacent to losers are flagged for removal at the start
of the next iteration.  Vertices with all losing edges are flagged as
winners.  At each iteration edges are removed from the graph, and
winning vertices are added to the MIS.  When the graph is empty, the
MIS is complete.

The first reduce() operation flags the two vertices in each edge as a
potential winner or loser.  This is done by comparing the two random
values for the vertices (or the randomized vertex IDs as explained
above).  The first collate() operation thus produces a multivalue for
each vertex $V_i$ which contains all the vertices it shares an edge
with, and associated winner/loser flags.  In the second reduce()
operation, if $V_i$ won the comparison with all its edge neighbors,
then it is a ``winner'' vertex.  All vertices connected to the winner
are emitted with a loser flag.  If $V_i$ is not an overall winner, its
edge vertices are emitted without the loser flag.  The second
collate() operation collects this new information for each vertex.

In the third reduce() operation, if $V_i$ was flagged as a loser by
any edge neighbor who was a winner, then it becomes a losing vertex
and emits all its edge vertices with a loser flag.  Otherwise it emits
all its edge vertices without the flag.

In the fourth reduce() operation, some vertices will have all their
edge vertices flagged as losers.  These vertices are all winners,
either the ones identified in the second reduce() operation, or
additional vertices who would become singletons (having no edges) when
edges containing a loser vertex are removed (on the next iteration).
All of these winning vertices are emitted to the MapReduce object that
stores the accumulating MIS.

All edges of each vertex are also emitted, retaining the loser flag if
they have it.  They are emitted with $E_ij$ as their key, with $V_i <
V_j$.  This is to insure both copies of the edge come together via the
fourth collate() operation.  The multivalue for each edge will now
have two values, either of which may be NULL or a loser flag.

At the first stage of the next iteration, any edge flagged by either
value as a loser does not compare its two vertices; it is effectively
deleted from the graph.  The Luby iterations end when all edges have
been removed from the graph.  At this point, the second MapReduce
object contains a complete MIS of vertices.
